Verifying Concurrent Search Structure Templates for C Programs using VST

### 2023-03-30 concurrency/CONCUR 2023/paper.tex
initial writing for CONCUR 2023 - wish us luck

\documentclass[a4paper,USenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

\bibliographystyle{plainurl}% the mandatory bibstyle

\usepackage{txfonts}
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{iris}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

\hyphenation{Comp-Cert}

\usepackage[utf8]{inputenc}

\title{Proving Atomic Specifications for Data Structures with Fine-Grained Locking}
\author{Roshan Sharma}{University of Illinois at Chicago, USA}{}{}{}
\author{Shengyi Wang}{Princeton University, USA}{}{}{}
\author{Alexander Oey}{, USA}{}{}{}
\author{Anastasiia Evdokimova}{University of Illinois at Chicago, USA}{}{}{}
\author{Lennart Beringer}{Princeton University, USA}{}{}{}
\author{William Mansky}{University of Illinois at Chicago, USA}{}{}{}

\authorrunning{R. Sharma, S. Wang, A. Oey, A. Evdokimova, L. Beringer, and W. Mansky} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Roshan Sharma, Shengyi Wang, Alexander Oey, Anastasiia Evdokimova, Lennart Beringer, and William Mansky} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\keywords{Dummy keyword} %TODO mandatory; please add comma-separated list of keywords

@@

\begin{abstract}
abstract
\end{abstract}

% Contributions:
% verified C implementation of BST -- the first?
% technique for hand-over-hand locking with share accounting -- slight modification of Gotsman's for logics without held predicates that remember the invariant; amounts to converting Locked_I into another share of the lock assertion
% technique for connecting locks with abstract state/atomicity
% VST-style decomposition of proofs into C-to-local-abstract and local-to-global-abstract
% lock-free implementation?

\section{Introduction}
\input{intro.tex}

\section{Background}
\label{background}
\subsection{VST and Iris}

We use the Verified Software Toolchain \cite{plfcc}  to prove the correctness of the concurrent binary search tree. VST is a Coq-based proof system that helps to prove separation logic properties of C programs with the assurance that the properties will be preserved in compiled code. First, we write programs in C programming and using CompCert’s \emph{clightgen}, the ASTs for C programs can be generated in Coq. We then state and prove specifications for them using Verifiable C. Verifiable C is a higher-order separation logic for reasoning about the functional correctness of C programs and carries a soundness proof that links high-level specifications to assembly code generated by the CompCert Verified compiler. Verifiable C supports reasoning about two separate kinds of concurrency:  Pthreads-style concurrency with locks (used by implementing blocking data structures that can be fine-grained or coarse-grained ), and concurrency using C11 atomic operations (i.e. lock-free data structures).

\ignore{\subsection{Ghost State and Global Invariants}
We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program. But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness. For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads. But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.   
\begin{figure}[htb]
\centering
$\texttt{x = 0;}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\qquad\pi = \pi_1\ .\ \pi_2$\\
$\begin{array}{l || l}
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{acquire(l);} & \texttt{acquire(l);}\\
\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{x++;} & \texttt{x++;}\\
\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{release(l);} & \texttt{release(l);}\\
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\end{array}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
\caption{The increment example annotated with separation logic assertion}
\label{figure1}
\end{figure}
We must preserve the connection between the value of shared resources and the work performed by each thread. A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread. They do not appear in the original program. Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}. In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together. VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} ($\gnamety$ in Coq) associated with the value $a$, which may be of any type. We will see different kind of ghost states used to verify binary search tree in the following sections.

The main idea behind the logic from Iris \cite{higherorderghoststate}, a mechnized higher-order concurrent separation logic framework, is the construction of \emph{global invariant} as \emph{ghost state}. \emph{Global Invariant} is an invariant on the global (ghost and physical) state of program. We can open any invariant but must close it again before taking any steps of execution unless those execution are atomic. Thread can use the contents of \emph{global invariant} during atomic operation if it can guarantee that no one will ever see an intermediate state in which invariant does not hold. The rules from Jung et al.\cite{higherorderghoststate} for creating and opening invariants are:

\begin{mathpar}
\inference[\textsf{inv\_alloc}]{}{\triangleright P \vdash \pvs[E] \mathsf{EX}\ i : \mathsf{iname}, \knowInv{i}{P}}

\inference[\textsf{inv\_open}]{i \in E}{\knowInv{i}{P} \vdash \pvs[E][E \setminus i] \triangleright P * (\triangleright P \wand \pvs[E \setminus i][E] \mathsf{emp})}
\end{mathpar}

where invariants are provided in the form of assertion $\knowInv{i}{P}$, and states that $P$ is maintained as an invariant on the global state with name $i$. The operator%%  $\pvs[][]$
is called ``fancy update" operator,that allows to allocate, open, and close the invariants and the later $\triangleright$ operator is used for impredicativity (i.e. $\knowInv{i}{..\knowInv{i}{P}..}$).}

\subsection{Hand-over-Hand Locking}
\label{selflock}

Hand-over-hand locking (also called lock coupling) is a fine-grained locking pattern for traversing a data structure in which we acquire the lock for the next node before releasing the lock for the current node, thus guaranteeing that the link between the two nodes is not removed or rearranged while we traverse it. Hand-over-hand locking is a classic example of a pattern that requires reference to lock invariants inside other lock invariants, dating back to the earliest papers on ``predicates in the heap'' in CSL~\cite{gotsman}. The trick is to assign each node in the data structure a lock invariant that includes the lock assertion for its child nodes; for a linked list, for instance, we might write $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock R(n')$, where $n'$ is the child of $n$ and $\ell'$ is $n'$'s lock. By acquiring $n$'s lock, we learn the lock assertion for $n'$, and can acquire its lock as needed before releasing $n$.

This invariant becomes slightly more complicated when we consider \emph{ownership} of locks, as is required if we want to eventually free the lock and return its resources. In this case the lock assertion carries a share $\pi$, where the full share $\pi = 1$ is required to free the lock, while any share is sufficient to acquire or release it. We can amend our lock invariant to $R(n) \triangleq n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n')$ for some suitable share $\pi$, but then, what happens to the share of $n_1$'s lock when we release the lock on $n$? Starting from a state in which we hold $n$'s lock $\ell$, this gives us:
\begin{align*}
&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') \}\\
&\texttt{acquire}(\ell');\\
&\{ \ell \islock_\pi R(n) * n \mapsto (d, n', \ell') * \ell' \islock_\pi R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}\\
&\texttt{release}(\ell);\\
&\{ \ell \islock_\pi R(n) * n' \mapsto (d, n'', \ell'') * \ell'' \islock_\pi R(n'') \}
\end{align*}
In order to re-establish $\ell$'s lock invariant $R(n)$, we must give up our knowledge of $n'$'s lock $\ell'$. This leaves us with two problems: we have no lock assertion for $\ell'$, so we will be unable to release it later, and we still have partial ownership of $\ell$'s lock assertion, without any way to re-collect our share. Both of these problems can be solved by making the lock \emph{recursive}, including a share of the lock assertion in its own lock invariant\footnote{Gotsman et al. instead use a separation assertion $\mathsf{Locked}$ that is always held by the thread that acquires a lock, and is required to free the lock: this is equivalent to an extra share of the lock assertion.}:
$$R(n) \triangleq \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n')$$

Share $\pi_1$ of each lock is held by the lock itself, while share $\pi_2$ is held by its parent. Then the same sequence of operations gives us:
\begin{align*}
&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') \}\\
&\texttt{acquire}(\ell');\\
&\{ \ell \islock_{\pi_1} R(n) * n \mapsto (d, n', \ell') * \ell' \islock_{\pi_2} R(n') * \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}\\
&\texttt{release}(\ell);\\
&\{ \ell' \islock_{\pi_1} R(n') * n' \mapsto (d, n'', \ell'') * \ell'' \islock_{\pi_2} R(n'') \}
\end{align*}
Now, when we acquire $\ell'$, we hold both $\pi_1$ and $\pi_2$ of its lock assertion; when we release $\ell$, we return our share $\pi_1$ of $\ell$ along with share $\pi_2$ of $\ell'$, but retain share $\pi_1$ of $\ell'$, so we can release it later. The details of defining a lock invariant that includes a share of its own lock assertion may vary across separation logics (VST provides a $\mathsf{selflock}$ assertion for exactly this purpose), but as long as it is possible, this pattern can be used to implement hand-over-hand lock invariants with share accounting, allowing us to traverse a data structure built with this pattern and then reclaim all of its resources once all threads are finished with it.

\subsection{Atomic Specifications}
\label{atomic}

For many concurrent data structures, the ideal correctness condition is that the data structure behaves the same as a sequential implementation, even when accessed simultaneously by multiple threads. This intuitive condition can be formalized as linearizability (operations appear to take effect in some total order) or atomicity (client threads never observe intermediate states of operations). In separation logic, atomicity can be expressed in the form of \emph{atomic triples}~\cite{tada}, written in the form $\forall a.\ \langle P_l\ |\ P_p(a)\rangle\ c\ \langle Q_l\ |\ Q_p(a)\rangle$, where $P_l$ and $Q_l$ are \emph{private} pre- and postconditions similar to an ordinary Hoare triple, and $P_p$ and $Q_p$ are \emph{public} pre- and postconditions, parameterized by an abstract value $a$ of the shared data structure. Intuitively, $P_l$ and $Q_l$ must be true before and after the call, while $P_p$ must be true for some value of $a$ at every point from the beginning of $c$ until some designated linearization point, at which point $Q_p$ becomes true atomically for the same value $a$. For instance, the specification $$\forall s.\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ s\rangle\ \texttt{push}(v)\ \langle \mathsf{is\_stack}\ p\ |\ \mathsf{stack}\ (v :: s)\rangle$$ expresses the fact that the \texttt{push} operation of a concurrent stack correctly implements the behavior of a sequential push, atomically transitioning from some stack $s$ to $v :: s$ at some point during its execution. %maybe something about linking ghost state, which parts are real vs. ghost, accessing invariants around atomic triple

VST has encoded an atomic Hoare triple, and provide the way that matches the notation VST uses for normal specifications. Such specification is called \emph{atomic specification}, and can be formalized in Coq as shown in  \ref{atomic-spec}.  This is how we write pre- and postcondition in VST. 
\begin{figure}[htb]
\centering
\begin{verbatim}
Program Definition insert_spec :=
DECLARE _insert
ATOMIC TYPE W OBJ a INVS Ei Eo
WITH ...
PRE [ ... ]
  PROP (...)
  LOCAL (...)
  SEP (P_l) | (P_p)
POST [ ... ]
  PROP ()
  LOCAL ()
  SEP (Q_l) | (Q_p)
\end{verbatim}
\caption{Atomic Specification in VST}
\label{atomic-spec}
\end{figure}
\texttt{W} is the \textsf{TypeTree} representing the type of arguments passed with \texttt{WITH} clause; \texttt{a} is the abstract state for the triple; \texttt{Ei} and \texttt{Eo} are the sets of invariants names inside and outside the triple. The \texttt{PROP} clause describes things that that are true independent of program state, the \texttt{LOCAL} clause describes the values contained in C local variables, and the \texttt{SEP} clause represents the \emph{separating conjunction} (*) of \emph{spatial predicates}, predicates on some part of the memory. In VST, while proving that a function implements an atomic specification, the precondition will contains an $\mathsf{atomic\_shift}$ assertion with the public pre- and postcondition, and the masks inside it. This atomic shift can be accessed through following two rules:
$$\inference[\textsf{atomic\_commit}]{\forall a, R * P_p\ a \Rrightarrow \mathsf{EX}\ y,\ Q_p\ a\ y * R'\ y}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$
$$\inference[\textsf{atomic\_rollback}]{\forall a, R * P_p\ a \Rrightarrow P_p\ a * R'}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R \Rrightarrow \textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * R'}$$
During commit, we must provide resources $R$ which, in combination with the public precondition $P_p$, allow us to prove the public postcondition $Q_p$. Then, we gain access to an assertion $Q$ required by the postcondition of the function, and leftover resources $R'$. During rollback, we provide resources $R$ which, in combination with the public precondition, allow us to reestablish the public precondition; we then regain the atomic shift back in the proof, as well as leftover resources $R'$. The rollback is specially used to learn some relationship between pieces of information (e.g. ghost state) stored in the public precondition, while the commit is used to prove the public postcondition from the precondition, and obtain an assertion $Q$. To complete the proof of any function's specification, we must always perform commit to obtain $Q$; we can perform any number of rollbacks before that point, but after that point we lose the atomic shift and no access to the public precondition.

\section{Locks and Atomicity}

\section{BST with Hand-over-Hand Locking}
\label{hand-over-hand}
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
\begin{mathpar}
\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{delete}(p, k)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{delete}\ t\ k)\rangle
\end{mathpar}
In our case, insert, lookup, and delete are simple Gallina functions on a functional tree data structure. The details of $\nodeboxrep$ and $\treerep$ may vary by implementation, but we expect that $\nodeboxrep$ contains at least a pointer to the data structure, and $\treerep$ contains some ghost state linking the concrete state of the data structure to the abstract tree $t$.

In the following sections, we describe the construction of these predicates, and the proofs of the operations, for BST implementations based on hand-over-hand locking and lock-free atomic operations.

%look at some of the Iris papers in CPP, etc. for how we should describe the proofs, e.g. section 4 of Formal Verification of a Concurrent Bounded Queue in a Weak Memory Model, Mevel and Jourdan, ICFP 2021

%%%%%%%%%%%%%%%%%%%%%

\ignore{In the previous section, we showed that how to prove concurrent data-structure is thread safe, but not functionally correct. To prove the correctness, our threads need to be able to record the information about the actions they have performed on the shared state, instead of sealing all knowledge of the shared data structure inside the lock invariant. We can accomplish this with ghost variables, a simple form of auxiliary state. Aside from the \emph{permission} on shared resources held by each thread that we used in the safety proofs, the verification of correctness properties also involves the \emph{ghost state}. In VST, any Coq type can be used as ghost state, as long as we can describe what happens when two elements of that type are joined together. We can use own predicate in VST to represent ghost state assertion as follows: $\mathsf{own}\ g\ a\ \mathit{pp}$, where $g$ is a \emph{ghost name}, $a$ is the value associated with $g$ which can be of any type, and $\mathit{pp}$ is a separation logic predicate.}

%Linearizability is one of the most popular strong correctness condition for the most of %concurrent data-structure, which implies that every operation appears to take place atomically, %in some order, consistent with the real-time ordering of those operations. 
%In order to record the action each thread performed as a linear history, we will define global %invariants using the ghost state, which are similar to lock invariants but are not associated %with any particular memory location. Instead, a global invariant is true before and after every %step of a program, acting as a publicly accessible resource. A program instruction can use the %contents of global invariant if it can guarantee that no one will ever see an intermediate state %in which the invariant does not hold.

\subsection{BST code}
\label{selflock}

\begin{figure}[htp]
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
typedef struct tree {int key; void *value; struct tree_t *left, *right;} tree;
typedef struct tree_t {tree *t; lock_t *lock;} tree_t;
typedef struct tree_t **treebox;
\end{lstlisting}
\begin{subfigure}[t]{0.5\textwidth}
\begin{lstlisting}[language = C, basicstyle=\small\ttfamily, numbers=left]
void insert (treebox t, int x, void *value) {
  struct tree_t *tgt = *t;
  struct tree *p;
  void *l = tgt->lock;
  acquire(l);
  for(;;) {
    p = tgt->t;
    if (p==NULL) {
      tree_t *p1 = malloc(sizeof *tgt);
      tree_t *p2 = malloc(sizeof *tgt);
      p1->t = NULL;
      p2->t = NULL;
      lock_t *l1 = malloc(sizeof(lock_t));
      makelock(l1);
      p1->lock = l1;
      release2(l1);
      lock_t *l2 = malloc(sizeof(lock_t));
      makelock(l2);
      p2->lock = l2;
      release(l2);
      p = malloc(sizeof *p);
      tgt->t = p;
      p->key=x; p->value=value; p->left=p1; p->right=p2;
      release(l);
      return;
    } else {
      int y = p->key;
      if (x<y){
      	tgt = p->left;
        void *l_old = l;
        l = tgt->lock;
        acquire(l);
        release(l_old);
      } else if (y<x){
        tgt = p->right;
        void *l_old = l;
        l = tgt->lock;
        acquire(l);
        release(l_old);
      } else {
      	p->value=value;
        release(l);
      	return;
      }
    }
  }
} 
\end{lstlisting} 
\end{subfigure}\qquad
\begin{subfigure}[t]{0.4\textwidth}
 \begin{lstlisting}[language = C, basicstyle=\small\ttfamily]
void *lookup (treebox t, int x) {
  struct tree *p; void *v;
  struct tree_t *tgt;
  tgt = *t;
  void *l = tgt->lock;
  acquire(l);
  p = tgt->t;
  while (p != NULL) {
    int y = p->key;
    if (x<y){
      tgt = p->left;
      void *l_old = l;
      l = tgt->lock;
      acquire(l);
      p = tgt->t;
      release(l_old);
    } else if (y<x){
      tgt = p->right;
      void *l_old = l;
      l = tgt->lock;
      acquire(l);
      p = tgt->t;
      release2(l_old);
    } else {
      v = p->value;
      release(l);
      return v;
    }
  }
  release(l);
  return NULL;
}
\end{lstlisting}
\end{subfigure}
\caption{BST with hand-over-hand locking}
\label{bst-conc}
\end{figure}

\subsection{Fine-grained locking and atomicity}
\label{atomicity}

To prove that an operation on a data structure satisfies a logically atomic specification, we must show that there are no visible intermediate states of the operation, i.e., that other threads see the data structure as unchanged until the linearization point at which the operation takes effect. In traditional concurrent separation logics such as VST's~\cite{oraclesematic}, this is done by protecting the data structure with a single lock whose \emph{lock invariant} describes consistent states of the data structure. In modern separation logics such as Iris~\cite{iris} or FCSL~\cite{fcsl}, we can make a series of atomic modifications to the data structure without continuously holding a lock, as long as one modification can be designated the \emph{linearization point} at which the new state becomes public, and all prior changes are invisible to other threads. Fine-grained locking, in which the pieces of a data structure are each protected by a separate lock with a traditional lock invariant, requires us to synthesize these two approaches: we may make a series of non-atomic modifications to a part of the data structure as long as we hold its lock, but when we release the lock we need to ensure that either we have reached the linearization point, or our changes are invisible to other threads.

We accomplish this using a technique derived from unpublished work by Jung~\cite{ralf-convo} for treating the critical sections of locks as atomic operations. The key is to include in each lock invariant a piece of ghost state that represents the abstract state of the locked section. From the outside, the effect of the critical section is to atomically update that piece of ghost state, and all other changes are internal to the lock. These pieces of ghost state can then be composed to yield the abstract state of the entire data structure. In effect, this construction turns a traditional lock invariant into a \emph{lock variant} $\ell \lockvar R$, where $R$ is parameterized by a value that can be updated in the critical section, as long as it corresponds to a linearization point or invisible change to the abstract state of the data structure as a whole. This allows traditional lock invariants to interact with atomic shifts and be used to prove atomic specifications.

The construction of the lock variants is simple. We begin with a \emph{part-reference} ghost state, a common construction for sharing information between an invariant and the threads that act on it, with elements $\publichalf{a}$ (the current state $a$) and $\myhalf{\pi}{p}$ (partial information about the current state, annotated with a share $\pi$). When the share $\pi$ is 1, we know that $p = a$. (do we want to define this ourselves, or reference it somewhere? note also that this is what Iris calls an authoritative algebra) A lock variant with an assertion $R(a)$ is then defined as %notation
$$\ell \lockvar_\pi R \triangleq \ell \islock \exists a.\ R(a) * \myhalf{\pi}{a}$$ %eliding ghost names here
The share $\pi$ may be set to 1 if the lock's information about $a$ is always fully up to date, or a smaller share if it can become outdated due to operations on other parts of the data structure. Since $a$ is existentially quantified, a thread that changes the value of $a$ forgets it after releasing the lock (as required in a lock invariant), but it must have synchronized with a $\publichalf{a}$ held elsewhere (usually in a global invariant) that will remember the change. Thus, the ghost state acts as a bridge between the component lock invariant and the global abstract state of the data structure. %smaller example before we get into BST?

From the atomic rules of section~\ref{atomic}, we derive the following rules for accessing and modifying pieces of the global abstract state:
\begin{mathpar}
\inference[\textsf{sync\_commit}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * \exists x_0',\,x_1'.\ (x_0, x_1) \leadsto (x_0', x_1')\ \land \\(\mathsf{my\_half}(g, \mathit{sh}, x_0') * \mathsf{public\_half}(g, x_1') \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}

\inference[\textsf{sync\_rollback}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{public\_half}(g, x_1) \Rrightarrow P_p\ a * R')}{\begin{array}{c}\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R  \Rrightarrow \\\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R'\end{array}}
\end{mathpar}
A thread that holds $\myhalf{}{a}$ (i.e., is in the critical section of a lock with a lock variant) may interact with the abstract state $P_p$ held in an $\textsf{atomic\_shift}$, as long as the corresponding $\publichalf{}$ is part of the abstract state. It may either leave both halves as is and gain updated resources $R'$ (including, e.g., a snapshot of the current abstract state), or update both halves to a new state that satisfies the public postcondition $Q_p$, executing the linearization point and obtaining the final postcondition $Q$. Typically we will use one of these rules at the end of each critical section, to show that the changes made in the critical section either are invisible to other threads (\textsf{sync\_rollback}) or have fulfilled the linearization point (\textsf{sync\_commit}). We can also derive simpler rules for special cases, such as a version of \textsf{sync\_commit} for read-only operations that do not change the state of any component:

$$\inference[\textsf{sync\_commit\_same}]{\forall a.\ R * P_p\ a \Rrightarrow \exists x_1.\ \mathsf{public\_half}(g, x_1) * (\mathsf{my\_half}(g, \mathit{sh}, x_0) * \mathsf{public\_half}(g, x_1) \Rrightarrow \exists y.\ Q_p\ a\ y * R'\ y)}{\textsf{atomic\_shift}(P_p, E_i, E_o, Q_p, Q) * \textsf{my\_half}(g,sh,x_0) * R \Rrightarrow \mathsf{EX}\ y,\ Q\ y * R'\ y}$$

\subsection{Specification of the hand-over-hand BST}
%rewrite to use the above: this is the global state made up of components, etc.

To verify our tree implementation, we need to instantiate $\nodeboxrep$ (the resources held by each client thread) and $\treerep$ (the abstract state of the data structure), and show how they interact to access and update the state of the data structure for each call. In a lock-based implementation, $\nodeboxrep$ will be a pointer to the lock of the root node, and acquiring that lock will give the thread access to more of the data structure. As described above, our approach is to associate a piece of ghost state with each node in the tree, which will be shared between that node's lock and the abstract state in $\treerep$. We also need to make sure that our lock invariants support the hand-over-hand pattern.

\subsubsection{Key Ranges}
When we traverse the BST looking for a key $k$, we will reach either a node containing $k$, or an empty node where $k$ would appear if it was in the tree. To prove correctness of this procedure, we need ghost state that tracks where each key ``should appear'' in the current tree. As observed by Krishna et al.~\cite{krishna2017flow}, this can be done by associating each node with a lower and upper bound on the keys appearing in the subtree rooted at that node. We refer to the product of these bounds as a \emph{range}. The range of a node is inherited from its parent, based on the parent's key: if node $n$ has range $(l, r)$ and key $k$, then its left child has range $(l, k)$ and its right child has range $(k, r)$. Figure~\ref{range_bst} shows an example of a BST with each node labeled with its range, starting with $(-\infty, +\infty)$ at the root and propagated to the empty \emph{leaf} nodes. 
\begin{figure}[htb]
\centering
\includegraphics[width=70mm,scale=0.5]{../FIG/range_prop.png}
\caption{An example of BST with each node labeled with \emph{range}; lower and upper bound for a key}
\label{range_bst}
\end{figure}
At the leaves, these ranges partition the space of all possible keys. If we reach a leaf node with range $(a, b)$, then we know that keys between $a$ and $b$ 1) are not currently in the tree and 2) can correctly be inserted at this leaf.  For example, suppose we want to insert a key $38$ into the tree in Figure~\ref{range_bst}. The right child of the node with key 35 is a leaf with range $(35, 40)$, so it is guaranteed to be the right place to insert 38.

>>>>

\documentclass[a4paper,UKenglish,cleveref, autoref, thm-restate]{lipics-v2021}
%This is a template for producing LIPIcs articles. 
%See lipics-v2021-authors-guidelines.pdf for further information.
%for A4 paper format use option "a4paper", for US-letter use option "letterpaper"
%for british hyphenation rules use option "UKenglish", for american hyphenation rules use option "USenglish"
%for section-numbered lemmas etc., use "numberwithinsect"
%for enabling cleveref support, use "cleveref"
%for enabling autoref support, use "autoref"
%for anonymousing the authors (e.g. for double-blind review), add "anonymous"
%for enabling thm-restate support, use "thm-restate"
%for enabling a two-column layout for the author/affilation part (only applicable for > 6 authors), use "authorcolumns"
%for producing a PDF according the PDF/A standard, add "pdfa"

@@

\bibliographystyle{plainurl}% the mandatory bibstyle
\usepackage{txfonts}
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
%% http://ctan.org/pkg/subcaption
\usepackage{uri}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{mathpartir}
\usepackage{semantic}
\usepackage{graphicx}
\usepackage{cases}
\usepackage{hyperref}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage{iris}
%\usepackage{lstlangcoq}
\usepackage[edges]{forest}
\renewcommand{\lstlistingname}{Figure}

@@

\title{Verifying Concurrent Search Structure Templates for C Programs using VST} %TODO Please add

%\titlerunning{Dummy short title} %TODO optional, please use if title is longer than one line

\author{Duc Than Nguyen}{University of Illinois at Chicago, USA \and \url{http://www.myhomepage.edu} }{johnqpublic@dummyuni.org}{https://orcid.org/0000-0002-1825-0097}{(Optional) author-specific funding acknowledgements}%TODO mandatory, please use full name; only 1 author per \author macro; first two parameters are mandatory, other parameters can be empty. Please provide at least the name of the affiliation and the country. The full address is optional. Use additional curly braces to indicate the correct name splitting when the last name consists of multiple name parts.

\author{William Mansky}{University of Illinois at Chicago, USA}{joanrpublic@dummycollege.org}{[orcid]}{[funding]}

\authorrunning{Duc Than Nguyen and William Mansky} %TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

\Copyright{Duc Than Nguyen and William Mansky} %TODO mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\keywords{...} %TODO mandatory; please add comma-separated list of keywords

\category{} %optional, e.g. invited paper

\relatedversion{} %optional, e.g. full version hosted on arXiv, HAL, or other respository/website
%\relatedversiondetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93]{Classification (e.g. Full Version, Extended Version, Previous Version}{URL to related version} %linktext and cite are optional

%\supplement{}%optional, e.g. related research data, source code, ... hosted on a repository like zenodo, figshare, GitHub, ...
%\supplementdetails[linktext={opt. text shown instead of the URL}, cite=DBLP:books/mk/GrayR93, subcategory={Description, Subcategory}, swhid={Software Heritage Identifier}]{General Classification (e.g. Software, Dataset, Model, ...)}{URL to related version} %linktext, cite, and subcategory are optional

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
... 
\end{abstract}

\section{Introduction}
\label{sec:introduction}
Krishna et al. proposed concurrent search structure templates~\cite{templates} as a method for separating the proof of correctness of a concurrent access method (optimistic concurrency, hand-over-hand locking, internal links) from the proof of correctness of the underlying data structure (linked list, hashtable, B-tree). Ideally, it should be possible to prove the correctness of $n$ (single-threaded) data structure implementations and $m$ concurrency patterns and obtain $n * m$ verified concurrent data structures. In practice, the story is more complicated: certain patterns only work for specific data structures or require the data structures to store extra information, while some internal data structure operations may not fit the template model. In this paper, we apply the template approach to concurrent data structure implementations in C, verified using the Verified Software Toolchain (VST)~\cite{plfcc}, and report on its effectiveness and the challenges we encountered. The template approach depends crucially on the idea of \emph{logically atomic specifications} introduced in TaDA~\cite{tada} and further developed in Iris~\cite{iris}; our proofs take advantage of recent work integrating Iris-style logical atomicity into VST~\cite{iris-vst-arxiv}.

\section{Background}
\subsection{Concurrent Search Structure Templates}

\subsection{Iris and VST}

\section{Search Structure Templates}
\subsection{Lock Coupling}

\subsection{Give-Up}

\subsection{Templates vs. Internal Reorganization}

\section{Verified Data Structures}
\subsection{Binary Search Tree}

%%
%% Bibliography
%%

%% Please use bibtex,

\bibliography{lipics}

\appendix

\section{Styles of lists, enumerations, and descriptions}\label{sec:itemStyles}

List of different predefined enumeration styles:

\end{document}
